Correcting for data truncation in Bayesian analyses

### 2017-05-25 papers/general_method/general.tex
Updating based on Alex's feedback

\begin{abstract}
Fainter objects are harder to see with a telescope than brighter ones. Lifetimes are strictly positive. We cut our data to remove background noise.  Truncated or biased datasets are inescapable in modern physics. In this paper, we present a simple overview of a Bayesian consideration of truncation, giving a solution to both analytically tractable and intractable models. This can be accomplished via a combination of analytic approximations and Monte Carlo integration, in which dataset simulation is efficiently used to correct for issues in the observed dataset. Toy models are included, along with numerical considerations and optimisations for implementation.
\end{abstract}

@@

This work provides a simple treatment of truncated data in a common Bayesian technique. Section \ref{sec:perfect} discusses the ever-elusive case of perfect data without truncation to provide a common basis for Sections \ref{sec:imperfect} and \ref{sec:real}, which respectively cover analytically correctable data truncation and analytically intractable models. Section \ref{sec:tricks} details numeric concerns and tricks to be aware of for effective implementation of Monte Carlo corrections applied to analytic approximations.

\section{Theory}

When formulating and fitting a model using a constraining dataset, we wish to resolve the posterior surface defined by
\begin{align}
P(\theta | {\rm data}) \propto P({\rm data} | \theta) P(\theta).
\end{align}
Of primary interest to us is the likelihood of observing the data given our parametrised model, $\mathcal{L} \equiv P({\rm data} | \theta)$. When dealing with experiments which have selection efficiency, our likelihood necessarily includes that efficiency, for we want to describe the probability that our observations were both drawn from the underlying theoretical model \textit{and} that those observations, given they happened, were subsequently successfully observed.  To make this extra conditional explicit, we can write our likelihood as
with selection effect $S$:
\begin{align}
\mathcal{L} &= P({\rm data} | \theta, S).
\end{align}
As we wish to describe our selection efficiency generally as a function dependent on both our data and our model, we can reformulate to
\begin{align}
\mathcal{L}&= \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{P(S|\theta)}.
\end{align}
Introducing an integral over all possible data to make $P(S|\theta)$ physical,
\begin{align}
\mathcal{L}&= \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{\int P(S, D|\theta)\, dD} \\
&= \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{\int P(S | D, \theta) P(D|\theta)\, dD}, \label{eq:main}
\end{align}
where the integral in the denominator has the same dimensionality is the experimental data. \textbf{Equation \ref{eq:main} is the generalised likelihood of experiments with sample selection.}

\section{The perfect world: No Truncation}
\label{sec:perfect}
In a perfect world, data is neither biased nor truncated. The data is perfect. Uncertainties are well quantified and normally distributed around true values. Presumably everything is also spherical and in a vacuum. Let us create a mock model in this perfect world. Let us observe a series of independent and identically distributed events $\vec{x}$, which are drawn from a normal distribution such that 
\begin{align}
\vec{x} \sim \mathcal{N}(\mu,\sigma).
\end{align}
If, having collected our observations $\vec{x}$, we wanted to constrain $\mu$ and $\sigma$, this would be a simple task of modelling the posterior surface. Taking uniform priors on both parameters, we simply wish to map the surface
\begin{align}
P(\theta | {\rm data}) \propto P({\rm data} | \theta) P(\theta),
\end{align}
where our model parameters $\theta = \lbrace \mu, \sigma \rbrace$ and our data is given by $\vec{x}$.
\begin{align}
P(\mu,\sigma| \vec{x}) &\propto P(\vec{x} | \mu, \sigma) P(\mu, \sigma)
\end{align}
With uniform priors, $P(\mu,\sigma) = {\rm constant}$, and can be absorbed into the constant of proportionality. Expanding our observation vector, the posterior surface is given by
\begin{align}
P(\mu,\sigma| \vec{x}) &\propto \prod_{i=1}^N \mathcal{N}(x_i | \mu, \sigma). \label{eq:prod}
\end{align}
Generating a hundred data points with $\mu=100,\ \sigma=10$, we can recover our input parameters easily, as shown in Figure \ref{fig:perfect}.
\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{example/perfect.pdf}
	\end{center}
	\caption{A systematic test of our perfect model, done by stacking the output chains from fitting 100 independent realisations of our 100 data points. Any systematic offset in our model would be revealed by a shift in the stacked results away from the true parameter values.}
	\label{fig:perfect}
\end{figure}

\section{The Imperfect World: Analytic Truncation}
\label{sec:imperfect}
In a slightly imperfect world we may have to deal with truncated data. For an example, consider the previous model, but with an instrumentation deficiency such that we can only observe events above a certain threshold, such that $x > \alpha$. We assign a value $\alpha=85$ for convenience. If we do not take this truncation into account, we will recover biased parameter estimates, as shown in Figure \ref{fig:imperfect}. However, we can correct for this truncation using equation \eqref{eq:main}.

In our example, our data is one dimensional (as we only observe one variable, $x$) and the selection efficiency is the step function $P(S|x,\theta) = \mathcal{H}(x - \alpha)$. Having successfully observed $x_i$, it follows that $x_i > \alpha$ and so $P(S|x_i,\theta) = 1$. To substitute in our normal model,
\begin{align}
\mathcal{L}_i &= \frac{P(S|x_i,\theta) P(x_i|\theta)}{\int P(S | D, \theta) P(D|\theta)\, dD} \\
&= \frac{ \mathcal{N}(x_i|\mu, \sigma)}{\int_{-\infty}^\infty \mathcal{H}(D - \alpha) \mathcal{N}(D|\mu, \sigma)\, dD} \\
&= \frac{ \mathcal{N}(x_i|\mu, \sigma)}{\int_{\alpha}^\infty \mathcal{N}(D|\mu, \sigma)\, dD} \\
&= \frac{ \mathcal{N}(x_i|\mu, \sigma)}{\frac{1}{2} {\rm erfc}\left[ \frac{\alpha - \mu}{\sqrt{2}\sigma} \right]}, 
\end{align}
where in the last line we have evaluated the integral in the case $\mu > \alpha$. Note that this is for a single observation, and so for a set of independent observations we need to introduce the product found in equation \eqref{eq:prod}. 
\begin{align}
\mathcal{L} &= \prod_{i=1}^N \mathcal{L}_i =  \prod_{i=1}^N \frac{ \mathcal{N}(x|\mu, \sigma)}{\frac{1}{2} {\rm erfc}\left[ \frac{\alpha - \mu}{\sqrt{2}\sigma} \right]}, 
\end{align}
However, as our selection efficiency correction is observationally independent it is identical for all observations, allowing us to take it outside the product.
\begin{align}
\mathcal{L} &= 2 \left( {\rm erfc}\left[ \frac{\alpha - \mu}{\sqrt{2}\sigma} \right] \right)^{-N}\prod_{i=1}^N  \mathcal{N}(x|\mu, \sigma), 
\end{align}
We can add this correction to our model, and note that we now recover unbiased parameter estimates, also shown in Figure \ref{fig:imperfect}.
\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{example/imperfect.pdf}
	\end{center}
	\caption{A systematic test of our imperfect model, done by stacking the output chains from fitting 100 independent realisations of our 100 data points, subject to our thresholding. The bias shown in the red `Biased' contour can be corrected to via the techniques shown in Section \ref{sec:imperfect} to recover unbiased surfaces.}
	\label{fig:imperfect}
\end{figure}

\section{The Real World: Analytically Intractable Truncation}
\label{sec:real}
Unfortunately it is a rare scenario when dealing with nature and all her faults for us to have an analytic selection function, let alone a function encapsulated by a single parameter. A more realistic scenario involves a selection efficiency instead would take the form of non-analytic function of many model parameters. And the function would probably be stochastic too, just to throw another wrench in the works. Provided a method of forward modelling or simulating observations, the solution is to combine an analytic approximate correction with Monte Carlo integration.

So let us modify our imperfect toy model. Instead of observing just one variable, $x$, we also observe a new independent variable, $y$, which is drawn from its own distribution $y \sim \mathcal{N}(\mu_y, \sigma_y)$. Our selection efficiency can now become a combination of $x$ and $y$, such that we only observe events that satisfy $x + \beta y > \alpha$, giving $P(S|x,y,\theta) = \mathcal{H}(x + \beta y - \alpha)$.  Our likelihood for such a toy model becomes now the combination of probabilities for observing both $x$ and $y$, with the denominator becoming an integral over all possible $X$ and $Y$ observations subject to our selection effects.
\begin{align}
\mathcal{L}_i &= \frac{ \mathcal{N}(x_i|\mu, \sigma) \mathcal{N}(y_i|\mu_y, \sigma_y)}
{\iint_{-\infty}^\infty \mathcal{H}(X + \beta Y - \alpha) \mathcal{N}(X|\mu, \sigma) \mathcal{N}(Y|\mu_y, \sigma_y)\, dX dY}
\end{align}
Assume that we cannot solve this integral analytically, and must resort to numeric solutions. These often clash with sampling methods, especially for high dimensional integrals. Inserting Monte Carlo integration into fitting algorithms can drastically slow them down, and algorithms such as Hamiltonian MCMC that require continuous surfaces can easily fail on surfaces that fluctuate from Monte Carlo integration. Even by fixing the samples used in MC integration (thereby giving a continuous surface), the complexity of the surface derivatives will pose almost insurmountable problems for any algorithms that utilise surface gradients. One solution is to find an approximate, analytic correction we can utilise in our fitting algorithm which seeks to shift the region of parameter space sampled by the sampler closer to the correct area.

@@

%\section*{Acknowledgments}

%We gratefully acknowledge the input of the many researchers that were consulted during the creation of this paper.

>>>>

\begin{abstract}
Lifetimes are strictly positive. We cut our data to remove background noise.  Astronomical observations suffer from Malmquist bias - our samples are preferentially biased toward those sources that are more likely to be observationally detected. These sample selections should be accounted for. In this paper we present a simple overview of a Bayesian consideration of sample selection, giving a solution to both analytically tractable and intractable models. This can be accomplished via a combination of analytic approximations and Monte Carlo integration, in which dataset simulation is efficiently used to correct for issues in the observed dataset. Toy models are included, along with numerical considerations and optimisations for implementation.
\end{abstract}

@@

This work provides a simple treatment of truncated data in a common Bayesian technique. The general theory for considering selection effects is discussed in Section \ref{sec:theory}. Section \ref{sec:examples} provides three examples of increasing complexity with sample selection. Section \ref{sec:tricks} details numeric concerns and tricks to be aware of for effective implementation of Monte Carlo corrections applied to analytic approximations.

\section{Theory}
\label{sec:theory}

When formulating and fitting a model using a constraining dataset, we wish to resolve the posterior surface defined by
\begin{align}
P(\theta | {\rm data}) \propto P({\rm data} | \theta) P(\theta).
\end{align}
Of primary interest to us is the likelihood of observing the data given our parametrised model, $\mathcal{L} \equiv P({\rm data} | \theta)$. When dealing with experiments which have selection efficiency, our likelihood necessarily includes that efficiency, for we want to describe the probability that our observations were both drawn from the underlying theoretical model \textit{and} that those observations, given they happened, were subsequently successfully observed.  To make this extra conditional explicit, we can write our likelihood as
with selection effect $S$:
\begin{align}
\mathcal{L} &= P({\rm data} | \theta, S). \label{eq:like}
\end{align}
As we wish to describe our selection efficiency generally as a function dependent on both our data and our model, we can reformulate to
\begin{align}
\mathcal{L}&= \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{P(S|\theta)}.
\end{align}
Introducing an integral over all possible data to make $P(S|\theta)$ physical,
\begin{align}
\mathcal{L}&= \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{\int P(S, D|\theta)\, dD} \\
&= \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{\int P(S | D, \theta) P(D|\theta)\, dD}, \label{eq:main}
\end{align}
where the integral in the denominator has the same dimensionality is the experimental data. \textbf{Equation \ref{eq:main} is the generalised likelihood of experiments with sample selection.}

\section{Sample Selection}
\label{sec:examples}

 In this section we present sample selections of increasing complexity. Each sample selection is accompanied by an illustrative example inspired by Type Ia supernova cosmology, where we characterize the properties of a standard or standardizable candle.

\subsection{Complete Selection}
\label{sec:perfect}
In a perfect world, data is neither biased nor truncated. The data is perfect. Uncertainties are well quantified and normally distributed around true values. Presumably everything is also spherical and in a vacuum. We thus begin by considering an ideal situation where the sample is complete.  All events are included in the sample such that per object, $P(S | {\rm data}, \theta)=1$.  Trivially this expression is independent of $\theta$, and our likelihood from equation \eqref{eq:main} reduces down to equation \eqref{eq:like}. As a concrete example of this case, let us consider a model for a population of objects whose brightnesses form a normal distribution with average $\mu$ and standard deviation $\sigma$.
Let us also assume that our experiment produces data $x$, measurements of the brightnesses with negligible measurement uncertainty, which we can formally state as
\begin{align}
\vec{x} \sim \mathcal{N}(\mu,\sigma).
\end{align}
If, having collected our observations $\vec{x}$, we wanted to constrain $\mu$ and $\sigma$, this would be a simple task of modelling the posterior surface. Taking uniform priors on both parameters, we simply wish to map the surface
\begin{align}
P(\theta | {\rm data}) \propto P({\rm data} | \theta) P(\theta),
\end{align}
where our model parameters $\theta = \lbrace \mu, \sigma \rbrace$ and our data is given by $\vec{x}$.
\begin{align}
P(\mu,\sigma| \vec{x}) &\propto P(\vec{x} | \mu, \sigma) P(\mu, \sigma)
\end{align}
With uniform priors, $P(\mu,\sigma) = {\rm constant}$, and can be absorbed into the constant of proportionality. Expanding our observation vector, the posterior surface is given by
\begin{align}
P(\mu,\sigma| \vec{x}) &\propto \prod_{i=1}^N \mathcal{N}(x_i | \mu, \sigma). \label{eq:prod}
\end{align}
Generating a hundred data points with $\mu=100,\ \sigma=10$, we can recover our input parameters easily, as shown in Figure \ref{fig:perfect}.
\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{example/perfect.pdf}
	\end{center}
	\caption{A systematic test of our perfect model, done by stacking the output chains from fitting 100 independent realisations of our 100 data points. Any systematic offset in our model would be revealed by a shift in the stacked results away from the true parameter values.}
	\label{fig:perfect}
\end{figure}

\subsection{Analytic Truncation}
\label{sec:imperfect}

We now consider a case with selection bias and has an analytic expression for the likelihood.  This illustrative case is useful for the reader as the influence of selection bias is simple and intuitive. The case we consider is identical to the one in the previous section, \textit{except} that only the subset of objects brighter than a threshold $\alpha$ can be observed.

With this sample selection, all events satisfy $x > \alpha$, giving $P(S|x,\theta) = \mathcal{H}(x - \alpha)$. We assign a value $\alpha=85$ for convenience. If we do not take this truncation into account, we will recover biased parameter estimates. However, we can correct for this truncation using equation \eqref{eq:main}, as the integral in the denominator has an analytic solution. Having successfully observed $x_i$, it follows that $x_i > \alpha$ and so $P(S|x_i,\theta) = 1$. To substitute in our normal model,
\begin{align}
\mathcal{L}_i &= \frac{P(S|x_i,\theta) P(x_i|\theta)}{\int P(S | D, \theta) P(D|\theta)\, dD} \\
&= \frac{ \mathcal{N}(x_i|\mu, \sigma)}{\int_{-\infty}^\infty \mathcal{H}(D - \alpha) \mathcal{N}(D|\mu, \sigma)\, dD} \\
&= \frac{ \mathcal{N}(x_i|\mu, \sigma)}{\int_{\alpha}^\infty \mathcal{N}(D|\mu, \sigma)\, dD} \\
&= \frac{ \mathcal{N}(x_i|\mu, \sigma)}{\frac{1}{2} {\rm erfc}\left[ \frac{\alpha - \mu}{\sqrt{2}\sigma} \right]}, 
\end{align}
where in the last line we have evaluated the integral in the case $\mu > \alpha$. Note that this is for a single observation, and so for a set of independent observations we need to introduce the product found in equation \eqref{eq:prod}. 
\begin{align}
\mathcal{L} &= \prod_{i=1}^N \mathcal{L}_i =  \prod_{i=1}^N \frac{ \mathcal{N}(x|\mu, \sigma)}{\frac{1}{2} {\rm erfc}\left[ \frac{\alpha - \mu}{\sqrt{2}\sigma} \right]}, 
\end{align}
However, as our selection efficiency correction is observationally independent it is identical for all observations, allowing us to take it outside the product.
\begin{align}
\mathcal{L} &= 2 \left( {\rm erfc}\left[ \frac{\alpha - \mu}{\sqrt{2}\sigma} \right] \right)^{-N}\prod_{i=1}^N  \mathcal{N}(x|\mu, \sigma), 
\end{align}
We can add this correction to our model, and note that we now recover unbiased parameter estimates. This is demonstrated in Figure \ref{fig:imperfect}, which shows the posterior surfaces for when you take sample selection into account and when you do not.
\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{example/imperfect.pdf}
	\end{center}
	\caption{A systematic test of our imperfect model, done by stacking the output chains from fitting 100 independent realisations of our 100 data points, subject to our thresholding. The bias shown in the red `Biased' contour can be corrected to via the techniques shown in Section \ref{sec:imperfect} to recover unbiased surfaces.}
	\label{fig:imperfect}
\end{figure}

\subsection{Analytically Intractable Truncation}
\label{sec:real}
Unfortunately it is a rare scenario when dealing with nature and all her faults for us to have an analytic selection function, let alone a function encapsulated by a single parameter. A more realistic scenario involves a selection efficiency instead would take the form of non-analytic function of many model parameters. And the function would probably be stochastic too, just to throw another wrench in the works. Provided a method of forward modelling or simulating observations, the solution is to combine an analytic approximate correction with Monte Carlo integration.

So let us modify our imperfect toy model from the previous section. Instead of observing just one variable, $x$, we also observe a new independent variable, $y$, which is drawn from its own distribution $y \sim \mathcal{N}(\mu_y, \sigma_y)$. Our selection efficiency can now become a combination of $x$ and $y$, such that we only observe events that satisfy $x + \beta y > \alpha$, giving $P(S|x,y,\theta) = \mathcal{H}(x + \beta y - \alpha)$.  Our likelihood for such a toy model becomes now the combination of probabilities for observing both $x$ and $y$, with the denominator becoming an integral over all possible $X$ and $Y$ observations subject to our selection effects.
\begin{align}
\mathcal{L}_i &= \frac{ \mathcal{N}(x_i|\mu, \sigma) \mathcal{N}(y_i|\mu_y, \sigma_y)}
{\iint_{-\infty}^\infty \mathcal{H}(X + \beta Y - \alpha) \mathcal{N}(X|\mu, \sigma) \mathcal{N}(Y|\mu_y, \sigma_y)\, dX dY}
\end{align}
Assume that we cannot solve this integral analytically, and must resort to numeric solutions. These often clash with sampling methods, especially for high dimensional integrals. Inserting Monte Carlo integration into fitting algorithms can drastically slow them down, and algorithms such as Hamiltonian MCMC that require continuous surfaces can easily fail on surfaces that fluctuate from Monte Carlo integration. Even by fixing the samples used in MC integration (thereby giving a continuous surface), the complexity of the surface derivatives will pose almost insurmountable problems for any algorithms that utilise surface gradients. One solution is to find an approximate, analytic correction we can utilise in our fitting algorithm which seeks to shift the region of parameter space sampled by the sampler closer to the correct area.

@@

\section*{Acknowledgments}

We gratefully acknowledge the input of the many researchers that were consulted during the creation of this paper.
